Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study (draft only, not to be quoted)

### 2020-06-13 pharmaco.Rmd
Updated to use a data.table object for case-control dataset

*Findings* -- In those without a diagnosis of a listed condition, severe COVID-19 was strongly associated with polypharmacy, defined by total number of drug classes dispensed: with those who had no dispensed prescriptions as reference group, the rate ratio (95% CIs) for COVID-19 associated with dispensing of 12 or more drug classes was XX.  In those with listed conditions, this association was in the reverse direction: the corresponding rate ratio was XX.  The regression coefficients for number of cardiovascular drug classes and number of other drug classes in a joint model were XX and XX respectively; thus the association with polypharmacy was specifically with non-cardiovascular drugs.  A backwards elimination procedure identified proton pump inhibitors, and analgesics (both opiates and non-opiate / compound preparations) as the drug classes with the largest independent contribution to the association of COVID-19 with polypharmacy.  Further analysis of the association with proton pump inhibitors showed a dose-response relationship with the average daily dose dispensed, that the association remained that the association was strongest in younger age groups: the rate ratio for COVID-19 associated with ever-use of a proton pump inhibitor varied from `r table.everuse.protonpump[3, 3]` in those aged 75 years and over to `r table.everuse.protonpump[1, 3]` in those aged under 60 years.

@@

Table \ref{tab:numdrugs.listed.any} shows that the COVID-19 was associated with polypharmacy in those without a listed condition, but that in those with a listed condition the direction of this association was reversed: those with dispensed prescriptions of 1 to 12 drug classes were at lower risk than those with no dispensed prescription.

The association of COVID-19 with polypharmacy was strongest in those who had no hospital diagnoses in the previous three years (because they had not been admitted to hospital).  In this group, with the no dispensed prescriptions category as baseline, the rate ratio ranged from `r table.numdrugsgr.num.icdch[2, 4]` in those with 1 to 3 drug classes dispensed to `r table.numdrugsgr.num.icdch[6, 4]` in those with more than 12 drug classes dispensed.

Risk of COVID-19 was strongly related to polypharmacy in all age bands: with no dispensed prescriptions as baseline category, the rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.

To study further the relationship with polypharmacy, we restricted the analysis to those not resident in a care home and without a diagnosis of a listed condition.  The rationale for this restriction was to reduce the frequency of factors including multi-morbidity and frailty which might be expected to confound the association between polypharmacy and COVID-19.  In people with listed conditions, we might expect prescribing of multiple drugs in accordance with evidence-based guidelines for management of these conditions.   Residents in care homes are at high risk of severe COVID-19 for reasons unrelated to polypharmacy, and the profile of drugs most commonly prescribed to people in care homes differs from the profile of drugs most commonly prescribed to people living independently.

## Associations of severe COVID-19 with drug classes
We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19.  Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.  These COVID-19-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.

A backward stepwise variable elimination procedure as described above was carried out to identify which specific drug classes contributed most to the association of COVID-19 with polypharmacy.  To save computation, the stepwise procedure was initiated with the subset of `r length(subparacols.forstepwisedrop)` drug classes in Table \ref{tab:subpara.coeffs} that showed univariate associations.  Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model.
Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.  The drug classes dropped at this stage included some of the most widely-prescribed drug classes that showed univariate associations with COVID-19 in Table \ref{tab:subpara.coeffs}, including lipid-lowering drugs, non-steroidal antiinflammatory agents and angiotensin-converting enzyme (ACE) inhibitors.  Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped. Cardiovascular drugs retained among these `r length(rows.print)` classes were loop diuretics and calcium channel blockers.  Four classes of drugs acting on the central nervous system were retained: other antidepressant drugs (excludes tricyclics and selective serotonin reuptake inhibitors), drugs used in nausea and vertigo, antipsychotic drugs, and non-opioid analgesics.  The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump, inhibitors, nonopioid /compound analgesics, and opioid analgesics.

## Associations of severe COVID-19 with proton pump inhibitor use
As the association of COVID-19 and exposure to proton pump inhibitors is not obviously explained by established co-morbidities, we explored this association in more depth.  As this examined a specific exposure, rather than polypharmacy which may have different profiles in care home residents or those diagnosed with listed conditions, the observations were not restricted to exclude care home residents or those diagnosed with a listed condition. Distinguishing between causality and confounding is challenging because proton pump inhibitors are usually co-prescribed with other drugs.  We used several approaches:

@@

* Stratification by propensity score for ever-use of a proton pump inhibitor.

>>>>

*Findings* -- In those without a diagnosis of a listed condition, severe COVID-19 was strongly associated with polypharmacy, defined by total number of drug classes dispensed. With those who had no dispensed prescriptions as reference group, the rate ratio (95% CI) for COVID-19 associated with dispensing of 12 or more drug classes was `r table.numdrugsgr.listed.any[6, 4]` in those without listed conditions.  In those with listed conditions, this association was in the reverse direction: the corresponding rate ratio was `r table.numdrugsgr.listed.any[12, 4]`.  The association was specifically with non-cardiovascular drugs: in a model with number of cardiovascular drug classes and number of other drug classes as covariates, the rate ratios were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`. 
A backwards elimination procedure identified proton pump inhibitors, and analgesics (both opiates and non-opiate / compound preparations) as the drug classes with the largest independent contribution to the association of COVID-19 with polypharmacy.  Further analysis of the association with proton pump inhibitors showed a dose-response relationship that was strongest in younger age groups: the rate ratio associated with average dispensed dose of 2 or more DDDs per day was `r table.dosegr.protonpump[5, 4]` in those under 60 years and `r table.dosegr.protonpump[15, 4]` in those aged under 60 years.

@@

Table \ref{tab:numdrugs.listed.any} shows that the COVID-19 was associated with polypharmacy in those without a listed condition, but that in those with a listed condition the direction of this association was reversed: those with dispensed prescriptions of 1 to 12 drug classes were at lower risk than those with no dispensed prescriptions. 
<!---
The association of COVID-19 with polypharmacy was strongest in those who had no hospital diagnoses in the previous three years (because they had not been admitted to hospital).  In this group, with the no dispensed prescriptions category as baseline, the rate ratio ranged from `r table.numdrugsgr.num.icdch[2, 4]` in those with 1 to 3 drug classes dispensed to `r table.numdrugsgr.num.icdch[6, 4]` in those with more than 12 drug classes dispensed.

Risk of COVID-19 was strongly related to polypharmacy in all age bands: with no dispensed prescriptions as baseline category, the rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over. 
--->

To study further the relationship with polypharmacy, we restricted the analysis to those without a diagnosis of a listed condition.  As prescribing of multiple drugs for primary and secondary prevention of cardiovascular disease is in accordance with evidence-based guidelines, we partitioned the number of drug classes dispensed between cardiovascular and other drugs.  In a model with number of cardiovascular drug classes and number of other drug classes as covariates, the rate ratios were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`.

## Associations of severe COVID-19 with drug classes
We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19 in those without listed conditions.  Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.  These COVID-19-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.

A backward stepwise variable elimination procedure as described above was carried out to identify which specific drug classes contributed most to the association of COVID-19 with polypharmacy.  To save computation, the stepwise procedure was initiated with the subset of `r length(subparacols.forstepwisedrop)` drug classes in Table \ref{tab:subpara.coeffs} that showed univariate associations.  Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model.
Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.  Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.  The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump, inhibitors, nonopioid /compound analgesics, and opioid analgesics.

The associations with these drug classes were examined in more depth.  For studies of these specific exposures the observations were not restricted to exclude those with listed conditions. 
 Distinguishing between causality and confounding is challenging because these drug classes  are usually co-prescribed with other drugs.  We used several approaches:

@@

* Stratification by propensity score.

## Associations of severe COVID-19 with proton pump inhibitor use
